---
title: "Reorganizing"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: "paged"
---

So far, we have always started out with neatly organized data files, and we have assumed that their contents are correct and in the form that we want. We won't always receive our data in the state that we would like. We might need to rework them a little before we can produce a plot or begin statistical analysis.

There are two packages that can help us to reorganize data. One of them, dplyr, we learned about in the last tutorial. The other is called `reshape2`. It is designed to make it easier to 'reshape', that is to switch between different alternative ways of organizing the same set of data.

```{r}
library(ggplot2)
library(dplyr)

library(reshape2)
```

As usual, we start by loading an example data set. We met these data briefly in the **Plotting** tutorial. Now we will look in a little more detail at their structure.

```{r}
salience = read.csv("data/salience.csv")
head(salience)
```

Unlike most of the other data sets we have seen so far, the **salience** data have a 'hierarchical' structure. The main outcome variables, **RT** and **Error** were measured mutiple times for each **Subject**, so the measurements are 'grouped' by subject. We will usually want to take this fact into account.
