---
title: "Summarizing"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: "paged"
---

We learned in the last tutorial how to display data in plots. A good plot can show what the data have to tell us, in a way that is clear and intuitive. However, one thing that plots can't do so well for us is to show exactly

Let's load again the **birth weights** data:

```{r}
bw = read.csv("data/birth_weights.csv")
head(bw)
```


We won't always receive our data in exactly the form we want them to have. We might need to rework the data a little before we can produce a plot or begin statistical analysis. As we have seen, many tasks in R can be performed more easily with the help of a package created specifically for that task. There are two important packages for working with data that we will learn about here: `reshape2` and `dplyr`. We will see what each of these can do below. First we must load them with `library()`.

We can see from the messages that the dplyr package 'masks' some basic R functions (we learned about masking in the **Packages** tutorial).

```{r}
library(ggplot2)

library(reshape2)
library(dplyr)
```

As usual, we start by loading an example data set. We met these data briefly in the **Plotting** tutorial. Now we will look in a little more detail at their structure.

```{r}
salience = read.csv("data/salience.csv")
head(salience)
```

Unlike most of the other data sets we have seen so far, the **salience** data have a 'hierarchical' structure. The main outcome variables, **RT** and **Error** were measured mutiple times for each **Subject**, so the measurements are 'grouped' by subject. We will usually want to take this fact into account.

* **Subject**: which subject was doing the task
* **Trial**: which attempt they were making at the task (first, second, third, etc.)
* **Orientation**: whether the object was unusually oriented
* **Luminance**: whether the object was unusually bright
* **SOA**: how long a delay elasped between preparing for the task and the object appearing (in ms)
* **RT**: the Reaction Time (in ms), i.e. how fast the person reacted
* **Error**: the amount of error made in the task, where a value of 0 indicates no error (i.e. perfect accuracy)




# The tidyverse

## ggplot


## dplyr


## reshape2



